---
title: "Assignment 1"
date: "`r Sys.Date()`"
author: "Johannes Hedström"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A66 Decision Theory \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Department of Computer and Information Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, fig.width = 5, fig.height = 3, fig.align = 'center', digits=10)

```

```{r}
library(knitr)
```


# Task 1

A bank official is concerned about the rate at which the bank’s tellers provide service
for their customers. He feels that all of the tellers work at about the same speed, which
is either 30, 40 or 50 customers per hour. Furthermore, 40 customers per hour is twice
as likely as each of the two other values, which are assumed to be equally likely. In
order to obtain more information, the official observes all five tellers for a two-hour
period, noting that 380 customers are served during that period.
Show that the posterior probabilities for the three possible speeds are approximately
0.000045, 0.99996 and 0.00000012 respectively.
(Hint: The total amount of 380 customers cannot be equalized with an average amount
of customers per teller and hour. Using formal calculations instead of simulating data, it
is easier to confirm the posterior probabilities.)


Prior probabilities:

$$P(30)=0.25, P(40)=0.5, P(50)=0.25$$

$$r = 380, t = \text{2 hours times 5 tellers} = 10$$

We get the posterior probabilities from bayes theorem when making inference on $\lambda$ in the poisson process:

$$P(\lambda \mid r, t)=\frac{\lambda^{r} e^{-t\cdot\ \lambda} \cdot P(\lambda)}{30^{r} e^{-t\cdot 30} \cdot 0.25+40^{r} e^{-t\cdot 40} \cdot 0.50+50^{r} e^{-t\cdot 50} \cdot 0.25}$$


Applying this on our data and lambda=30 would look like this:

$$P(\lambda=30 \mid r=380, t=10)=\frac{30^{380} e^{-10\cdot\ 30} \cdot 0.25}{30^{380} e^{-10\cdot 30} \cdot 0.25+40^{380} e^{-10\cdot 40} \cdot 0.50+50^{380} e^{-10\cdot 50} \cdot 0.25}$$

This will create problems when calculating as $30^{380}$ will be infinity, so instead we take the log of the formula which gives us the following: 

Likelihood * Prior:
$$log(\lambda^{r} e^{-t\cdot\ \lambda} \cdot 0.25)= r\cdot log(\lambda)-t\lambda - log(4) $$

The full expression is then: 

\begin{align} \log (P(\lambda \mid r, t))=r \log (\lambda)-t \lambda-\log(4) - \\ 
\left ((r \ log(30) -10 \cdot 30 - log(4)) + (r \ log(40) -10 \cdot 40 - log(2)) + (r  log(50) -10 \cdot 50 - log(4) \right)) \end{align}

These calculations needs to be scaled to be able to get the exponent for the logged values, this can be done by subtracting with the largest value as calculating exp on large negative values is not a computational problem.

This results in : 

$$L(30,480,10)\cdot P(30) =  380\cdot log(30)-10 \cdot 30 - log(4) = 991$$

$$L(40,480,10)\cdot P(40) = r\cdot log(40)-10\cdot 40 - log(2) = 1001$$

$$L(50,480,10)\cdot P(50) = r\cdot log(\lambda)- 10 \cdot 50 - log(4) = 985 $$
When subtracting with max we get:

$$ LP_{30} - Max(LP)= 991-1001= -10 $$

$$LP_{40} - Max(LP)= 1001-1001 = 0 $$

$$LP_{50} - Max(LP) = 985-1001 = - 16$$

We can now anti log the values.

$$ exp(-10) \approx 4.484337e-05$$
$$ exp(0) = 1 $$
$$ exp(-16) \approx 1.245451e-07$$

Now we normalize these values by the sum of them:


$$P(\lambda=30 \mid r=380, t=10) = \frac{4.484337e-05}{ 4.484337e-05 +1 + 1.245451e-07} \approx 0.000045 $$
$$P(\lambda=40 \mid r=380, t=10) = \frac{1}{ 4.484337e-05 +1 +1.245451e-07} \approx 0.99996 $$
$$P(\lambda=50 \mid r=380, t=10) = \frac{ 1.245451e-07} { 4.484337e-05 +1 + 1.245451e-07} \approx 0.00000012 $$


Below we have a function which checks that this will return the correct results

```{r}
log_P <- function(lambda, r, t,p) {
  # First part of the expression: r * log(lambda) - t * lambda - log(4)
  part1 <- r * log(lambda) - t * lambda + log(p)
  
  # Subracting with maximum of the log values to rescale values
  part1 <- exp(part1-max(part1))
  
  # Final result: Normalizing the values with sum
  log_P_value <- part1 / sum(part1)
  
  return(log_P_value)
}

lambda <- c(30,40,50)
r <- 380
t <- 10
p <- c(0.25,0.5,0.25)

# running the function
res <- data.frame(log_P(lambda, r,t,p))

colnames(res) <- 'Posterior probabilities'

rownames(res) <- c('P(30)', 'P(40)', 'P(50)')

knitr::kable(res, digits=10)

```


Another way of solving this question could be with utilizing the exponential class of distributions.

$$f(x|\theta) = f(x|\lambda) = e^{(ln\lambda)\cdot x - lnx! -\lambda}$$
Where $A(\lambda)= ln \lambda\cdot t$,$B(x)=380$, $-lnx!=-\Gamma(x+1)=C=-\Gamma(381)$(constant) and $D(\lambda)=-\lambda\cdot t$.

This results in :

$$ f(x|30) = e^{(ln300)\cdot 380 - \Gamma 381 -300}$$

$$f(x|40) = e^{(ln 400)\cdot 380 - \Gamma 381 -400}$$

$$f(x|50) = e^{(ln 500)\cdot 380 - \Gamma 381 -500}$$

Then from these results use your prior and likelihood to get the posterior probabilities. 


$$P(\lambda=30|data) = \frac{f(x|30)\cdot0.25}{f(x|30)\cdot0.25+ f(x|40)\cdot0.50+ f(x|50)\cdot0.25}$$

$$P(\lambda=40|data) = \frac{f(x|40)\cdot0.50}{f(x|30)\cdot0.25+ f(x|40)\cdot0.50+ f(x|50)\cdot0.25}$$

$$P(\lambda=50|data) = \frac{f(x|50)\cdot0.25}{f(x|30)\cdot0.25+ f(x|40)\cdot0.50+ f(x|50)\cdot0.25}$$


```{r}

# calculating likelihood
lik <- exp(log(t*lambda)*r-lfactorial(r+1)-t*lambda)

# posterior probabilities
posteriot<- lik*p/(sum(lik*p))

res <- data.frame(posteriot)

colnames(res) <- 'Posterior probabilities'

rownames(res) <- c('P(30)', 'P(40)', 'P(50)')

knitr::kable(res, digits=10)

```

Same results as in the question and as the first method, and this is the preferred way to solve the question as its a much simpler.


# Task 2

Assume you have decided to bet on a horse race, and that you have very little
knowledge about the competing horses. You consider betting on Little Joe, and you see
that the odds for this horse are 9 to 1 (i.e. odds against that the horse will win). You
decide to look up some historical tracks on how Little Joe has performed recently and
note that he has won in 2 of the last 10 races he competed in. You can assume that these
races are fairly comparable with respect to the levels of his competitors.

* a) Using the above as your background information, what are your subjective odds for
Little Joe?

* b) If you bet, you will obtain 9 times the money you have put in the bet. What is your
subjective expected return from betting on Little Joe?


## a)



$$P_E = P(E|I,n,E) = \frac{a+n_E}{a+b+n}$$


$$P_E = P(E|I,n,E) = \frac{1+2}{1+9+10} = \frac{3}{20}= 0.15$$

$$Odds_E = \frac{P_E}{1-P_E}$$

$$Odds_E = \frac{0.15}{0.85} = 3/17 = 0.1764706 $$
 
 
 
```{r}
Pe <- 3/20 # prob 

Odds_pe <- Pe/(1-Pe) # odds

3/17 == Odds_pe # check if the same

```


My subjective odds is that the odds against that Little Joe will win is 17 to 3 and the odds for Little Joe is then 3 to 17.




## b)

Subjective expected return 

$$P_E \cdot 9 + (1-P_E) \cdot {-1}$$

Using our subjective probabilities gives us: 

$$0.15 \cdot 9 + 0.85 \cdot {-1} = 0.5$$

```{r}
# Expected Return

(9*0.15) - 0.85

```
 
The subjective expected return from the bet is 0.5, which mean that on average you expect to make a 50% profit for every money spent on the bet. 


