---
title: "Assignment 3"
date: "`r Sys.Date()`"
author: "Johannes Hedström"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A66 Decision Theory \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Department of Computer and Information Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, fig.width = 5, fig.height = 3, fig.align = 'center', digits=10)

```

```{r, echo=FALSE}
library(knitr)
```



**Prerequisites and facts:**

* You will travel from City A to Airport B for catching a flight on a specific day

* You need to be at Airport B at latest 9:30 a.m. to not miss the flight

* You can choose between taking the train or going by car (no other means of conveyance)

* The train journey takes 3 hours according to the schedule. It departs from City A at 6 a.m. (and arrives at Airport B at 9 a.m. if on schedule)

* Going by car takes approximately 2 hours (assume it to be exactly 2 hours for simplicity), but you need to add 15 minutes for parking your car. The car is available from 6 a.m.

* The train ticket costs 50 €

* The total cost for using the car (parking included) is 70 €

**Potential obstacles and costs; assumptions and assigned prior probabilities:**

* The probability that the train is delayed by x minutes is (45 – x)0.001

* It can be assumed that the train will not derail or break down

* The additional cost from missing your flight is 300 €

* The probability of encountering an unexpected traffic incident/jam that delays your journey by y minutes is (90 – y)0.0002 conditional on that the car is not involved in an accident

* It can be assumed that the car will not break down

* The probability of the car (and you) being involved in an accident is 0.01

* If the car is involved in an accident you will not catch your flight 
















