---
title: "Assignment 4"
date: "`r Sys.Date()`"
author: "Johannes Hedström"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A66 Decision Theory \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Department of Computer and Information Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, fig.width = 7, fig.height = 3, fig.align = 'center', digits=10)

```

```{r, echo=FALSE}
library(knitr)
```


# Question 1

Assume some budget calculations depend on whether a certain cost will be at least SEK
120 000 or lower than this amount. A reasonable model for this cost is a normal
distribution with standard deviation SEK 12 000 (independent of the mean) and a mean that can be modeled as normally distributed with mean 115 000 (SEK) and standard
deviation 9 000 (SEK). No trend is anticipated for this cost and for the 6 previous
periods the average cost was SEK 121 000.
Note that the hypotheses are about the actual cost, not the expected cost.


## a 

Show that the prior odds for the hypothesis that the cost will exceed SEK 120 000
(against the alternative that it will not) is about 0.59. [hint: write the observed variableas a sum of two independent random variable $\tilde x = \tilde \mu + \tilde \epsilon$]


$$\tilde \mu \sim N (115000, 9000)$$

Standard deviation of $\tilde x$ should then be 

$$\sqrt{ 9000^2 + 12000^2} = 15000$$

Now you have the mean and the standard deviation so you can calculate the standardized value for 120000.

$$z = \frac{120000- 115000}{15000}$$

From the normal cumulative distribution(standard) we get that the probability for x to be less than 120 000 to be 0.63056:

so the prior odds to exceed 120000 is then 

$$\frac{1-0.63056}{0.63056} \approx 0.59$$


```{r}
sd <- sqrt(9000^2+12000^2 )

p <- pnorm(120000,115000,sd) # normal distribution mean of x< 120000

p <- round(p,5)

paste('The probability for the cost to be less than 120 0000 is', p)

paste('The prior odds for the cost to exceed 120 000 is', round(((1-p)/p),3)) # odds

```

rounds to 0.59.


## b

Show that the Bayes factor (considering the average cost for the previous 6 periods)
for the hypothesis that the cost will exceed SEK 120 000 (against the alternative that
it will not) is about 1.63 [Not about 1.60 or about 1.70].


$$B = \frac{Odds(H_0|data,I)}{Odds(H_A|data,I)}= \frac{P(H_0|data,I)/P(H_A|data,I)}{P(H_0|I)/P(H_A|I)}$$

$$H_A:\text{The cost is lower or equal to 120000}$$

$$H_0:\text{The cost is exceeds 120000}$$

Prior odds for $H_0|I$ is calculated in a.


The standard deviation when the data is collected is then 

$$\sqrt{\frac{\sigma^2}{6}} = \frac{15000}{\sqrt{6}} = 6123.724$$
We had a mean of 121000 for the collected data so our likelihood then follows:

$$\tilde x \sim N (121000, 6123.724^2)$$





```{r}

datamean <- 121000
dsd <- 15000 / sqrt(6)  

z_data <- (120000 - dmean) / dsd


H0 <- 1 - pnorm(z_data)  # P(cost > 120000 | data)
HA <- pnorm(z_data)      # P(cost <= 120000 | data)


bayes_factor <- (H0/HA)/((1-p)/p)

bayes_factor
```




## c

If the loss of accepting the hypothesis that the cost will be lower than SEK 120 000 while the opposite will be true is SEK 4 000, and the loss of accepting the
hypothesis that the cost will be at least SEK 120 000 while the opposite will be true is SEK 6 000, which decision should be made for the budget (according to the rule of minimizing the expected loss)?


# Question 2

Consider a big box filled with an enormous amount of poker chips. You know that either 70% of the chips are red and the remainder blue, or 70% are blue and the remainder red. You must guess whether the big box has 70% red / 30% blue or 70% blue / 30% red. If you guess correctly, you win US dollar 5 . If you guess incorrectly, you lose US $ 3 . Your prior probability that the big box contains 70% red / 30% blue is 0.40, and you are risk neutral in your decision making (i.e. your utility is linear in money).




## a 

If you could purchase sample information in the form of one draw of a chip from the big box, how much should you be willing to pay for it? Assume now that the cost of sampling is US$0.25 (i.e. 25 US cents) per draw.

```{r}
d_matrix <- data.frame('actions'=c('Guess 70 % red', 'Guess  70 % blue'), 'State 1(70%red)'=c(5,-3), 'State 2(70%blue)'=c(-3,5))

colnames(d_matrix) = c('actions','State 1(70%red)','State 2(70%blue)')
knitr::kable(d_matrix)
```

Expected utility:

$$\text{Action red} = 5\cdot 0.4 -3\cdot 0.6 = 0.2$$

$$\text{Action blue} = -3\cdot 0.4 + 5\cdot 0.6 =1.8 $$
Posterior distribution:

$$P(\theta|BUY(sample)) = \frac{P(BUY|\theta)\cdot P(\theta)}{P(BUY)}$$

Where $\theta$ is the states.

Probability of drawing a red chip then (blue is then the opposite):

$$P(Red) = 0.7\cdot0.4 + 0.3\cdot0.6 = 0.46$$

So when the state is red then:
$$P(State 1|Red)=\frac{0.7\cdot0.4}{0.46} = 0.6087$$

$$P(State 2|Red)= 1-P(State 1|Red) = 0.3913$$


Expected utility after sample

Guessing 70% red (after observing red):
$$E[U(a^{'})∣Red]=5 \cdot 0.6087 - 3\cdot 0.3913 = 1.8696$$

Guessing 70% blue (after observing red):
$$E[U(a^{''})∣Red] -35 \cdot 0.6087 +5 \cdot 0.3913 = 0.1304$$


When utility is linear in money, we could also define VSI as

$$ VSI(y_n) = E(R(a^{''}|y_n)|y_n) - E(R(a^{'})|y_n) $$

So the VSI in this case is:

$$1.8696 - 1.8 = 0.0696$$
We dont have to calculate for when drawing a blue chip as the cost is symmetric, so the results will be the same. 





Expected value of sample information(EVSI)


$$EVSI= \sum_y VSI(y_n)\cdot f(y_n)dy_n$$

$$ EVSI = 0.46 \cdot 0.0696 + 0.54⋅0.0696 = 0.1392  $$

The amount you should be willing to pay is 0.1392.



## b

What is the ENGS for a sample of 10 chips using a single-stage sampling plan.